---
title: "stephanie_analysis"
output: html_document
date: "2024-04-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(tidyr)
library(leaps)
library(randomForest)
library(datasets)
library(caret)
library(glmnet)
library(ggpubr)
library(car)
library(FSA)
library(ggstatsplot)
```

# Data Cleaning

## Upload

```{r}
data <- read.csv("All_Data_UVA_Deidentified_04-08-24.csv",header=T)
data_math <- read.csv("Math_Dataset_UVA_Deidentified_03-29-24.csv",header=T)
data_read <- read.csv("Reading_Dataset_UVA_Deidentified_02-02-24.csv",header=T) 
```

## Clean

```{r}
data_read$Actual.Achievement <- as.integer(data_read$Actual.Achievement) 
data_read$Growth <- as.integer(data_read$Value.Added.Score) 
data_math$Actual.Achievement <- as.integer(data_math$Actual.Achievement)
data_math$Growth <- as.integer(data_math$Growth)
data$Actual.Achievement <- as.integer(data$Actual.Achievement)
data$Growth <- as.integer(data$Growth)

data_r <- data_read %>% 
  mutate(Type.of.Teacher.Certification = fct_explicit_na(Type.of.Teacher.Certification, na_level = "NONE"))
data_r <- data_r %>% 
  mutate(Type.of.Principal.Certification = fct_explicit_na(Type.of.Principal.Certification, na_level = "NONE"))
data_r$Year.of.Teacher.Certification <- factor(data_r$Year.of.Teacher.Certification)
data_r <- data_r %>% 
  mutate(Year.of.Teacher.Certification = fct_explicit_na(Year.of.Teacher.Certification, na_level = "Not applicable"))
data_r$Type.of.Teacher.Certification <- factor(data_r$Type.of.Teacher.Certification,levels=c("NONE","MICRO","FULL"))

data_m <- data_math %>% 
  mutate(Type.of.Certification = fct_explicit_na(Type.of.Certification, na_level = "NONE"))
data_m <- data_m %>% 
  mutate(Type.of.Principal.Certification = fct_explicit_na(Type.of.Principal.Certification, na_level = "NONE"))
data_m$Year.of.Certification <- factor(data_m$Year.of.Certification)
data_m <- data_m %>% 
  mutate(Year.of.Certification = fct_explicit_na(Year.of.Certification, na_level = "Not applicable"))
data_m$Type.of.Certification <- factor(data_m$Type.of.Certification,levels=c("NONE","MICRO","FULL"))


data_a <- data %>% 
  mutate(Type.of.Certification = fct_explicit_na(Type.of.Certification, na_level = "NONE"))
data_a <- data_a %>% 
  mutate(Type.of.Principal.Certification = fct_explicit_na(Type.of.Principal.Certification, na_level = "NONE"))
data_a$Year.of.Certification <- factor(data_a$Year.of.Certification)
data_a <- data_a %>% 
  mutate(Year.of.Certification = fct_explicit_na(Year.of.Certification, na_level = "Not applicable"))
data_a$Type.of.Certification <- factor(data_a$Type.of.Certification,levels=c("NONE","MICRO","FULL"))
```

# Analysis

## Teacher Certification Only - Growth

### Reading

#### Regression

```{r}
lm_teacher_cert_r <- lm(data_r$Growth~ data_r$Type.of.Teacher.Certification) 
summary(lm_teacher_cert_r)
```

#### ANOVA - Assumption Tests

```{r}
one.way <- aov(Growth ~ Type.of.Teacher.Certification, data = data_r)

# are equal variances met?
lev <- leveneTest(Growth ~ Type.of.Teacher.Certification, data = data_r)
summary(lev)
lev

plot(one.way,1)

# is normalcy met?
plot(one.way,2)

resids <- residuals(one.way)
shapiro.test(resids)
```

Normal distribution not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Growth ~ Type.of.Teacher.Certification, data = data_r)
```

There are statistically significant differences. To determine which categories are different from each other, Dunn's Test is conducted.

```{r}
dunnTest(Growth ~ Type.of.Teacher.Certification, data = data_r, method="holm")
```

```{r}
data_r$Type.of.Teacher.Certification <- factor(data_r$Type.of.Teacher.Certification,levels=c("FULL","MICRO","NONE"))
ggbetweenstats(data=data_r,
               x= Type.of.Teacher.Certification,
               y=Growth,
               title = "Effect of Teacher Certification on Growth",
               subtitle = "Reading Dataset",
               results.subtitle = FALSE,
               type = "nonparametric",
               plot.type = "box",
               pairwise.comparisons = TRUE,
               pairwise.display = "significant",
               centrality.plotting = FALSE,
               bf.message = FALSE,
               ylab="Growth",
               xlab="Teacher Certification Type")
```

### Mathematics

#### Regression

```{r}
lm_teacher_cert_m <- lm(data_m$Growth~ data_m$Type.of.Certification) 
summary(lm_teacher_cert_m)
```

#### ANOVA - Assumption Tests

```{r}
one.way_m <- aov(Growth ~ Type.of.Certification, data = data_m)  
# are equal variances met? 
lev_m <- leveneTest(Growth ~ Type.of.Certification, data = data_m) 
summary(lev_m) 
lev_m  
plot(one.way_m,1)  
# is normalcy met? 
plot(one.way_m,2)  
resids_m <- residuals(one.way_m) 
shapiro.test(resids_m)
```

Normal distribution not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Growth ~ Type.of.Certification, data = data_m)
```

There are statistically significant differences. To determine which categories are different from each other, Dunn's Test is conducted.

```{r}
dunnTest(Growth ~ Type.of.Certification, data = data_m, method="holm")
```

```{r}
data_m$Type.of.Teacher.Certification <- factor(data_m$Type.of.Certification,levels=c("FULL","MICRO","NONE")) 

ggbetweenstats(data=data_m,                
               x= Type.of.Certification,                
               y=Growth,                
               title = "Effect of Teacher Certification on Growth",                
               subtitle = "Reading Dataset",                
               results.subtitle = FALSE,                
               type = "nonparametric",                
               plot.type = "box",                
               pairwise.comparisons = TRUE,                
               pairwise.display = "significant",                
               centrality.plotting = FALSE,                
               bf.message = FALSE,                
               ylab="Growth",                
               xlab="Teacher Certification Type")
```

### Full Dataset

#### Regression

```{r}
lm_teacher_cert_a <- lm(data_a$Growth~ data_a$Type.of.Certification)  
summary(lm_teacher_cert_a)
```

#### ANOVA - Assumption Tests

```{r}
one.way_a <- aov(Growth ~ Type.of.Certification, data = data_a)   

# are equal variances met?  
lev_a <- leveneTest(Growth ~ Type.of.Certification, data = data_a)  
summary(lev_a)  
lev_a   
plot(one.way_a,1)   
# is normalcy met?  

plot(one.way_a,2)   
resids_a <- residuals(one.way_a)  
shapiro.test(resids_a)
```

Normal distribution not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Growth ~ Type.of.Certification, data = data_a)
```

There are statistically significant differences. To determine which categories are different from each other, Dunn's Test is conducted.

```{r}
dunnTest(Growth ~ Type.of.Certification, data = data_a, method="holm")
```

```{r}
data_a$Type.of.Certification <- factor(data_a$Type.of.Certification,levels=c("FULL","MICRO","NONE"))   
ggbetweenstats(data=data_a,                                
               x= Type.of.Certification,                                
               y=Growth,                                
               title = "Effect of Teacher Certification on Growth",               
               subtitle = "Combined Dataset",                                
               results.subtitle = FALSE,                                
               type = "nonparametric",                                
               plot.type = "box",                                
               pairwise.comparisons = TRUE,                                
               pairwise.display = "significant",                                
               centrality.plotting = FALSE,                                
               bf.message = FALSE,                                
               ylab="Growth",                                
               xlab="Teacher Certification Type")
```

## Teacher Certification Only - Actual Achievement

### Reading

#### Regression

```{r}
lm_teacher_cert_ra <- lm(data_r$Actual.Achievement~ data_r$Type.of.Teacher.Certification)  
summary(lm_teacher_cert_ra)
```

#### ANOVA - Assumption Tests

```{r}
one.way_ra <- aov(Actual.Achievement ~ Type.of.Teacher.Certification, data = data_r)  
# are equal variances met? 
lev_ra <- leveneTest(Actual.Achievement ~ Type.of.Teacher.Certification, data = data_r) 
summary(lev_ra) 
lev_ra  
plot(one.way_ra,1)  
# is normalcy met? 
plot(one.way_ra,2)  
resids_ra <- residuals(one.way_ra) 
shapiro.test(resids_ra)
```

Normal distribution not met, equal variance assumptions not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Actual.Achievement ~ Type.of.Teacher.Certification, data = data_r)
```

There are no statistically significant differences.

```{r}
dunnTest(Actual.Achievement ~ Type.of.Teacher.Certification, data = data_r, method="holm")
```

```{r}
data_r$Type.of.Teacher.Certification <- factor(data_r$Type.of.Teacher.Certification,levels=c("FULL","MICRO","NONE")) 
ggbetweenstats(data=data_r,                
               x= Type.of.Teacher.Certification,                
               y=Actual.Achievement,                
               title = "Effect of Teacher Certification on Actual Achievement",                
               subtitle = "Reading Dataset",                
               results.subtitle = FALSE,                
               type = "nonparametric",                
               plot.type = "box",                
               pairwise.comparisons = TRUE,                
               pairwise.display = "significant",                
               centrality.plotting = FALSE,                
               bf.message = FALSE,                
               ylab="Actual Achievement",                
               xlab="Teacher Certification Type")
```

### Mathematics

#### Regression

```{r}
lm_teacher_cert_ma <- lm(data_m$Actual.Achievement~ data_m$Type.of.Certification)  
summary(lm_teacher_cert_ma)
```

#### ANOVA - Assumption Tests

```{r}
one.way_ma <- aov(Actual.Achievement ~ Type.of.Certification, data = data_m)   
# are equal variances met?  
lev_ma <- leveneTest(Actual.Achievement ~ Type.of.Certification, data = data_m)  
summary(lev_ma)  
lev_ma   
plot(one.way_ma,1)   
# is normalcy met?  
plot(one.way_ma,2)   
resids_ma <- residuals(one.way_ma)  
shapiro.test(resids_ma)
```

Normal distribution not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Actual.Achievement ~ Type.of.Certification, data = data_m)
```

There are statistically significant differences. To determine which categories are different from each other, Dunn's Test is conducted.

```{r}
dunnTest(Actual.Achievement ~ Type.of.Certification, data = data_m, method="holm")
```

```{r}
data_m$Type.of.Teacher.Certification <- factor(data_m$Type.of.Certification,levels=c("FULL","MICRO","NONE"))   
ggbetweenstats(data=data_m,                                
               x= Type.of.Certification,                                
               y=Actual.Achievement,                                
               title = "Effect of Teacher Certification on Actual Achievement",                                
               subtitle = "Math Dataset",                                
               results.subtitle = FALSE,                                
               type = "nonparametric",                                
               plot.type = "box",                                
               pairwise.comparisons = TRUE,                                
               pairwise.display = "significant",                                
               centrality.plotting = FALSE,                                
               bf.message = FALSE,                                
               ylab="Actual Achievement",                                
               xlab="Teacher Certification Type")
```

### Full Dataset

#### Regression

```{r}
lm_teacher_cert_aa <- lm(data_a$Actual.Achievement~ data_a$Type.of.Certification)   
summary(lm_teacher_cert_aa)
```

#### ANOVA - Assumption Tests

```{r}
one.way_aa <- aov(Actual.Achievement ~ Type.of.Certification, data = data_a)     

# are equal variances met?   
lev_aa <- leveneTest(Actual.Achievement ~ Type.of.Certification, data = data_a)   
summary(lev_aa)   
lev_aa    
plot(one.way_aa,1)    

# is normalcy met?    
plot(one.way_aa,2)    
resids_aa <- residuals(one.way_aa)   
shapiro.test(resids_aa)
```

Normal distribution not met.

#### Kruskal-Wallis Test

```{r}
kruskal.test(Actual.Achievement ~ Type.of.Certification, data = data_a)
```

There are statistically significant differences. To determine which categories are different from each other, Dunn's Test is conducted.

```{r}
dunnTest(Actual.Achievement ~ Type.of.Certification, data = data_a, method="holm")
```

```{r}
data_a$Type.of.Certification <- factor(data_a$Type.of.Certification,levels=c("FULL","MICRO","NONE"))    
ggbetweenstats(data=data_a,                                                
               x= Type.of.Certification,                                                
               y=Actual.Achievement,                                                
               title = "Effect of Teacher Certification on Actual Achievement",                               
               subtitle = "Combined Dataset",                                                
               results.subtitle = FALSE,                                                
               type = "nonparametric",                                                
               plot.type = "box",                                                
               pairwise.comparisons = TRUE,                                                
               pairwise.display = "significant",                                                
               centrality.plotting = FALSE,                                                
               bf.message = FALSE,                                                
               ylab="Actual Achievement",                                                
               xlab="Teacher Certification Type")
```

# All Predictors

## Growth

```{r}
reduced_a <- select(data_a,-TestTest,-Test.Name.,-Generated.Unique.Teacher.ID,-Expected.Achievement,-Actual.Achievement,-Vertical.Scaled.Score.,-Actual.Achievement.Percentile,-Compacted.Math,-Grade)
reduced_aa <- select(data_a,-TestTest,-Test.Name.,-Generated.Unique.Teacher.ID,-Expected.Achievement,-Growth,-Vertical.Scaled.Score.,-Actual.Achievement.Percentile,-Compacted.Math,-Grade)

reduced_m <- select(data_m,Race,Gender,English.Learner,Students.with.Disabilities,Grade,Certified.Principal,Type.of.Principal.Certification,Teacher.License.Type,Teacher.Years.Experience,Teacher.Highest.Degree,Teacher.Birth.Year,Teacher.Gender,Teacher.Race.Ethnicity,Certified.Teacher,Type.of.Certification,Year.of.Certification,Growth)
reduced_ma <- select(data_m,Race,Gender,English.Learner,Students.with.Disabilities,Grade,Certified.Principal,Type.of.Principal.Certification,Teacher.License.Type,Teacher.Years.Experience,Teacher.Highest.Degree,Teacher.Birth.Year,Teacher.Gender,Teacher.Race.Ethnicity,Certified.Teacher,Type.of.Certification,Year.of.Certification,Actual.Achievement)

reduced_r <- select(data_r,-Test.Name.,-Generated.Unique.Teacher.ID,-Expected.Achievement,-Actual.Achievement,-Vertical.Scaled.Score.,-Actual.Achievement.Percentile,-Grade, -Generated.Unique.Student.ID, -Value.Added.Score)
reduced_ra <- select(data_r,-Test.Name.,-Generated.Unique.Teacher.ID,-Expected.Achievement,-Growth,-Vertical.Scaled.Score.,-Actual.Achievement.Percentile,-Grade, -Generated.Unique.Student.ID, -Value.Added.Score)
```

### Stepwise

#### All

```{r}
reduced_a_nona <- na.omit(reduced_a)
regnull_a <- lm(Growth ~ 1, data = reduced_a_nona)
regfull_a <- lm(Growth~., data = reduced_a_nona)
stepwise_lm_a <- step(regnull_a, scope = list(lower= regnull_a, upper=regfull_a), direction = "both")
```

```{r}
summary(stepwise_lm_a)
```

#### Reading

```{r}
reduced_r_nona <- na.omit(reduced_r)
regnull_r <- lm(Growth ~ 1, data = reduced_r_nona)
regfull_r <- lm(Growth~., data = reduced_r_nona)
stepwise_lm_r <- step(regnull_r, scope = list(lower= regnull_r, upper=regfull_r), direction = "both")
```

```{r}
summary(stepwise_lm_r)
```

#### Math

```{r}
reduced_m_nona <- na.omit(reduced_m)
regnull_m <- lm(Growth ~ 1, data = reduced_m_nona)
regfull_m <- lm(Growth~., data = reduced_m_nona)
stepwise_lm_m <- step(regnull_m, scope = list(lower= regnull_m, upper=regfull_m), direction = "both")
```

```{r}
summary(stepwise_lm_m)
```
